A highly-available move operation for replicated trees

### 2021-02-10 paper/review-response.tex

\documentclass[10pt]{article}
\usepackage[a4paper,margin=1in]{geometry}
\usepackage{spverbatim}
\sloppy

\usepackage{xr}
\externaldocument{move-op}

\newcommand{\todo}[1]{\textcolor{red}{TODO: #1}}
\newcommand{\authorcomment}[1]{\begin{quote}\textbf{Author comment:} #1\end{quote}}

\begin{document}
\title{Summary of changes of ``A highly-available move operation for replicated trees''}
\author{Martin Kleppmann, Dominic P.\ Mulligan, Victor B.F.\ Gomes, Alastair R.\ Beresford}
\date{}
\maketitle

We thank the reviewers very much for their comments, which have improved the paper.
In this document we explain how we have addressed the issues highlighted in the reviews.
The most important changes are as follows:
\begin{itemize}
\item We have changed the title (previously \emph{``A highly-available move operation for replicated trees and distributed filesystems''}) to remove \emph{``and distributed filesystems''}, as suggested by Reviewer 2.
\end{itemize}

\section{Reviewer 1}

\begin{spverbatim}
Recommendation: Author Should Prepare A Major Revision For A Second Review

Comments:
Introduction section offers very brief and simplified coverage of common topics and don't really offer anything new.
It would be even more valuable if some critical comparisons were made among the systems and approaches. There is no discussion that could help the reader understand that, given a particular system/approach configuration or set of performance or availability goals, approach X would be preferable to approach Y.
In Section 3.5, please explain all cases with examples.
My main concern is whether the comparison is fair. It shows that some results are achievable. How could one trustfully decide to apply the operations on the basis of these results for another scenario?

Additional Questions:
1.  Please explain how this manuscript advances this field of research and/or contributes something new to the literature.: It presents move operation for distributed filesystems.

2. Is the manuscript technically sound? Please explain your answer under Public Comments below.: Yes

1. Which category describes this manuscript?: Practice / Application / Case Study / Experience Report

2. How relevant is this manuscript to the readers of this periodical? Please explain your rating under Public Comments below.: Relevant

1. Are the title, abstract, and keywords appropriate? Please explain under Public Comments below.: Yes

2. Does the manuscript contain sufficient and appropriate references? Please explain under Public Comments below.: Important references are missing; more references are needed

3. Does the introduction state the objectives of the manuscript in terms that encourage the reader to read on? Please explain your answer under Public Comments below.: Could be improved

4. How would you rate the organization of the manuscript? Is it focused? Is the length appropriate for the topic? Please explain under Public Comments below.: Could be improved

5. Please rate the readability of the manuscript. Explain your rating under Public Comments below.: Readable - but requires some effort to understand

6. Should the supplemental material be included? (Click on the Supplementary Files icon to view files): Yes, as part of the digital library for this submission if accepted

7. If yes to 6, should it be accepted: After revisions.  Please include explanation under Public Comments below.

8. Would you recommend adding the code/data associated with this paper to help address your concerns and/or strengthen the paper?: Yes

Please rate the manuscript. Please explain your choice.: Good
\end{spverbatim}

\section{Reviewer 2}

\begin{spverbatim}
Recommendation: Author Should Prepare A Major Revision For A Second Review

Comments:
## Is the manuscript technically sound?
Yes. As stated before the paper has a clear contribution, which includes a mechanised proof of the algorithm correctness. Moreover, the proof code is presented in the appendix and is available online with instructions on how to check proofs locally.

## How relevant is this manuscript to the readers of this periodical?
The topic is relevant for TPDS community as it proposes a distributed algorithm to support optimistic replication of tree data structures.

## Are the title, abstract, and keywords appropriate?
Yes. However, it is unclear why is necessary to include "distributed filesystems" in the title. The work is concerned with a tree replicated datatype and distributed filesystems is an application.

## Does the manuscript contain sufficient and appropriate references?
The following reference should be included and is missing:

De Porre K, Myter F, De Troyer C, Scholliers C, De Meuter W, Boix EG. Putting order in strong eventual consistency. In IFIP International Conference on Distributed Applications and Interoperable Systems 2019 Jun 17 (pp. 36-56). Springer, Cham.

De Porre et al. work proposes an optimistic replication algorithm that can transform a general-purpose data type into a replicated data type. Similarly to the submitted paper, strong eventual consistency is ensured by maintaining a global total order of the operations. Since the case study in De Porre's paper is an AVL tree, a comparison should be included in the related work.

## Does the introduction state the objectives of the manuscript in terms that encourage the reader to read on?
I found the introduction enticing and, although brief, it does a great job in defining the context, illustrating the problem, and clearly defining the contributions.

## How would you rate the organization of the manuscript? Is it focused? Is the length appropriate for the topic?
The paper is well structured. I particularly liked section 2 that clearly illustrates the main issues with concurrent non-coordinated move operations. Also, by showing that the replicated filesystems of Dropbox and Google drive both have bugs demonstrates how difficult is to devise a correct resolution policy for the move operation. Overall, section 2 nicely shows the work relevance and potential impact without getting too technical.
The replication algorithm is thoroughly described and the proof of correctness is presented at the right level of detail - although the proofs are an essential component a more detailed presentation would become tedious.

## Please rate the readability of the manuscript.
The paper is well written and easy to follow.

## Detailed comments

C1: The system model and network requirements should be explicitly defined.

C2: Is the metadata parameter necessary? It seems that this parameter could be abstracted away which would simplify the algorithm presentation.

C3: Further justification is needed for choosing to encode add and remove operations as move operations. My intuition, base on previously published results as CISE and Hamsaz, is that not all operations need to follow a total ordered across replicas. However, because all operations are defined as moves, conflicts are always possible. I was left wondering if by having explicit add and remove operations some undo-do-redo cycles could be avoided.

C4: Instead of having an undo-do-redo cycle would it be possible to have state checkpoints? As operations might arrive in any order, the number of operations to undo can be high, in that case, it might be useful to have state checkpoints and combine these with undo operations.

C5: Notation is not consistent. It is stated that types are primed letters (2nd paragraph of section 3.1), however, this naming convention is dropped later on.

C6: The evaluation proves the point that the tree CRDT is the best option for fast response time and disconnected mode (in this case it is the only option).

C7: The choice of using the extracted code for the evaluation is questionable. Stating that the code could be optimised, does not provide useful information. It is unclear what the impact would be if optimisations were applied. It's reasonable to conclude that the comparison between the proposed CRDT and state machine replication would be maintained, but the optimised code would give a clear evaluation of the tree CRDT.

## Overall evaluation

I really enjoyed reading this paper, it's well written and easy to follow and addresses a relevant topic to the TPDS community.
It has a novel contribution as it defines an optimistic replication algorithm for trees that ensures convergence and the structural properties of trees. The idea behind the tree CRDT is simple, but it provides a solution to an existing problem. The fact that authors found problems in widely used applications that could be addressed with the tree CRDT proves its usefulness.
The fact the CRDT was formally verified for any number of replicas and a certified implementation was extracted from the formal definitions, strengths considerably the submission.
There are some choices that need to be further discussed and the evaluation should consider an optimised implementation of the algorithm.

Additional Questions:
1.  Please explain how this manuscript advances this field of research and/or contributes something new to the literature.: The paper presents the first tree replicated data type that allows concurrent move operations without any coordination between replicas. The tree data type relies on an optimistic replication algorithm that ensures state convergence and that the tree structure properties are preserved at all times. Given the complex (inductive) structure of trees, a strong point of the work is the mechanised proof that provides certification for the correctness of the algorithm.

2. Is the manuscript technically sound? Please explain your answer under Public Comments below.: Yes

1. Which category describes this manuscript?: Research/Technology

2. How relevant is this manuscript to the readers of this periodical? Please explain your rating under Public Comments below.: Relevant

1. Are the title, abstract, and keywords appropriate? Please explain under Public Comments below.: Yes

2. Does the manuscript contain sufficient and appropriate references? Please explain under Public Comments below.: Important references are missing; more references are needed

3. Does the introduction state the objectives of the manuscript in terms that encourage the reader to read on? Please explain your answer under Public Comments below.: Yes

4. How would you rate the organization of the manuscript? Is it focused? Is the length appropriate for the topic? Please explain under Public Comments below.: Satisfactory

5. Please rate the readability of the manuscript. Explain your rating under Public Comments below.: Easy to read

6. Should the supplemental material be included? (Click on the Supplementary Files icon to view files): Yes, as part of the digital library for this submission if accepted

7. If yes to 6, should it be accepted: After revisions.  Please include explanation under Public Comments below.

8. Would you recommend adding the code/data associated with this paper to help address your concerns and/or strengthen the paper?: No

Please rate the manuscript. Please explain your choice.: Excellent
\end{spverbatim}

\section{Reviewer 3}

\begin{spverbatim}
Recommendation: Accept With No Changes

Comments:
The paper presents an implementation of a replicated tree data type with operations for adding, removing, and moving tree nodes. Replicated tree data types are essential basic blocks of a wide range of distributed software like distributed file systems, editors, applications manipulating XML/JSON structures, etc. While replication is needed to ensure resilience to network/machine failures, an important and difficult challenge in this context is ensuring some form of convergence, e.g., that all the nodes storing a replica of the data structure (tree) reach the same state when all the updates have been propagated in the network. This is an instance of strong eventual consistency. The paper presents a solution to this challenge that provides high availability as well (clients do not need to wait for the replicas to synchronize), and that is also formally verified in the Isabelle/HOL theorem prover.

The paper starts with a very clear description of the motivation, noting that current solutions for the above challenge, deployed in Google Drive or Dropbox, are not satisfactory. The Google Drive solution may lead to unexpected errors while the Dropbox solution may duplicate nodes of the tree which may impact the user experience. It identifies the main challenge in designing replicated trees which is related to the move operation: moving a tree node so that it becomes the child of a different parent. In fact, node insertion and removal can be reduced to move operations.

Then, it describes an implementation that in particular, solves the issues in Google Drive or Dropbox, which is based on a very neat and simple idea. Essentially, every operation is associated with a timestamp (one can use Lamport timestamps [11] and avoid synchronized clocks), and every replica maintains a tree and a log of operations (issued locally or remote operations received from other replicas), ordered by timestamps. Applying a remote operation O with timestamp T consists in undoing operations with a timestamp bigger than T, applying O, and finally, re-applying the undone operations. This solution, which is quite generic and could be applied for other data types as well, makes it possible to handle concurrent move operations that try to move the same node (taking into account only the one with a bigger timestamp like in Last Writer Wins conflict resolution [10]), and avoid introducing cycles due to concurrent move operations. It also ensures strong eventual consistency because when all the updates are received, any two replicas apply them in the same order (according to their timestamps). The fact that the implementation satisfies strong eventual consistency and that every replica stores a tree (without cycles) is proved in the Isabelle/HOL theorem prover. The implementation in Isabelle is compiled to executable Scala code which is evaluated in a rather realistic deployment (with three replicas on different continents) and compared to a standard master-slave implementation where the order between operations is managed by a single master node. This evaluation shows that indeed, the current solution has very low latency compared to the master-slave implementation, although a smaller throughput.

The related work is quite extensive, covering the literature on Conflict-Free Data Types (the implementation in this paper being an instance of such a datatype), distributed file systems (the main application of this work), and works that use similar implementation ideas, based on timestamps.

Overall, I think that the paper is very well written. The motivation and the description of the challenges are very convincing, the flow of the ideas being very easy to follow. Technically, the implementation relies on an idea which is simple in hindsight, but which has not been used in previous work (up to my knowledge), and which seems to have a very important practical impact. The formalization and proofs in Isabelle/HOL are a very important and non-trivial contribution which is meant to increase the confidence we have in this solution.

Additional Questions:
1.  Please explain how this manuscript advances this field of research and/or contributes something new to the literature.: The paper presents a new formally-verified implementation of a replicated tree data type, which is an essential ingredient of a wide range of distributed software like distributed file systems, editors, applications manipulating XML/JSON structures, etc. This implementation is highly-available (the latency of applying an operation is independent of the network's latency) and eventually consistent (all replicas reach the same state when all updates have been propagated). These properties are essential in modern distributed services. The implementation solves existing issues in industrial applications like Google Drive and Dropbox, and contradicts claims about the existence of such an implementation in previous research articles.

2. Is the manuscript technically sound? Please explain your answer under Public Comments below.: Yes

1. Which category describes this manuscript?: Research/Technology

2. How relevant is this manuscript to the readers of this periodical? Please explain your rating under Public Comments below.: Very Relevant

1. Are the title, abstract, and keywords appropriate? Please explain under Public Comments below.: Yes

2. Does the manuscript contain sufficient and appropriate references? Please explain under Public Comments below.: References are sufficient and appropriate

3. Does the introduction state the objectives of the manuscript in terms that encourage the reader to read on? Please explain your answer under Public Comments below.: Yes

4. How would you rate the organization of the manuscript? Is it focused? Is the length appropriate for the topic? Please explain under Public Comments below.: Satisfactory

5. Please rate the readability of the manuscript. Explain your rating under Public Comments below.: Easy to read

6. Should the supplemental material be included? (Click on the Supplementary Files icon to view files): Yes, as part of the digital library for this submission if accepted

7. If yes to 6, should it be accepted: As is

8. Would you recommend adding the code/data associated with this paper to help address your concerns and/or strengthen the paper?: Yes

Please rate the manuscript. Please explain your choice.: Excellent
\end{spverbatim}
\end{document}

### 2021-02-26 paper/move-op.tex
New data processing scripts and graphs
Update paper title
Address reviewer comments

\begin{document}
\title{A highly-available move operation for replicated trees and distributed filesystems}

@@

In distributed filesystems and collaborative multi-user software, this tree is replicated across multiple nodes.
If users attempt to update the tree concurrently on different replicas, concurrency control is required.
However, standard techniques such as two-phase locking require synchronous coordination between replicas.
If we want the system to continue processing read and write requests even in the presence of arbitrary network partitions (in other words, if we want \emph{availability} and \emph{partition-tolerance} in the sense of the CAP theorem~\cite{Gilbert:2002il}), then we must avoid such coordination.
This is an important practical consideration since network partitions are common in many networks~\cite{Alquraan:2018wj,Bailis:2014jx}.

We can use \emph{optimistic replication}~\cite{Saito:2005jw}, which means that any replica can make changes to the data without waiting for communication with any other replicas; operations are sent to other replicas at some later time when a network connection is available.
This approach has the advantage of enabling disconnected operation on mobile devices, and high availability in geo-replicated settings.
It also improves performance for end users because the duration of a user request is independent of any network latency.

@@

Our algorithm supports optimistic replication, allowing replicas to temporarily diverge as they are updated, and ensuring that they always converge towards a consistent state.
It is an example of a \emph{Conflict-free Replicated Data Type} or CRDT~\cite{Shapiro:2011un}, and it guarantees a consistency model called \emph{strong eventual consistency}~\cite{Shapiro:2011un,Gomes:2017gy}.
Our contributions are:
\begin{itemize}
    \item We define a Conflict-free Replicated Data Type for trees that allow \emph{move} operations without any coordination between replicas such as locking or consensus.
        As discussed in \S\ref{sec:impossibility}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
    \item We formalise the algorithm using Isabelle/HOL \cite{DBLP:conf/tphol/WenzelPN08}, a proof assistant based on higher-order logic, and obtain a computer-checked proof of correctness.
        In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge to a consistent state, while preserving the tree structure.
        Our proof technique can also be applied to other distributed systems, making it of independent interest.
    \item To demonstrate the practical viability of our approach, we refine the algorithm to an executable implementation within Isabelle/HOL and prove the equivalence of the two.
        We extract a formally verified Scala implementation from Isabelle and evaluate its performance with replicas across three continents.
    \item We perform experiments with Dropbox and Google Drive, and show that they exhibit problems that would be prevented by our algorithm.
\end{itemize}

@@

On a filesystem, the destination directory of a move operation must not be a subdirectory of the directory being moved.
For example, if \texttt{b} is a subdirectory of \texttt{a}, then the Unix shell command \texttt{mv a a/b/} will fail with an error.
This restriction is required because allowing this operation would introduce a cycle into the directory graph, and so the filesystem would no longer be a tree.
This restriction is required in any other tree structure that supports a move operation.

@@

Another possible resolution is shown in Figure~\ref{fig:move-cycle}b: the nodes involved in the concurrent moves (and their children) could be duplicated, so that both ``$A$ as a child of $B$'' and ``$B$ as a child of $A$'' can exist in the tree.
However, such duplication is undesirable for the same reasons as in \S\ref{sec:move-same-item}.

@@

\section{The replicated move operation}\label{sec:algorithm}

@@

When the user wants to make a change to the tree, they generate an operation and apply it to their local replica.
Every operation is also asynchronously sent over the network to all other replicas, and applied to every remote replica using the same algorithm as for local operations.
The network may deliver operations in any order; thus, the communication can be performed peer-to-peer, and it does not require any central server or consensus protocol.
We only assume that eventually every operation is applied on every replica, provided that network partitions are of a finite duration.

The key consistency property of our algorithm is \emph{convergence}: that is, whenever any two replicas have applied the same set of operations, then they must be in the same state---even if the operations were applied in a different order on different replicas.
We prove this in \S\ref{sec:proof} by showing that applying operations is commutative.

@@

We allow the tree to be updated in three ways: by creating a new child of any parent node, by deleting a node, or by moving a node to be a child of a new parent.
However, all three types of update can be represented by a move operation.
To create a node, we generate a fresh ID for that node, and issue an operation to move this new ID to be a child of its desired parent; the node is then implicitly created.
We also designate as ``trash'' some node ID that does not exist in the tree; then we can delete a node by moving it to be a child of the trash.

@@

When a replica applies a \isa{Move} operation to its tree, it also records a corresponding \isa{LogMove} operation in its log.
The \isa{t}, \isa{p}, \isa{m} and \isa{c} fields are taken directly from the \isa{Move} record, while the \isa{oldp} field is filled in based on the state of the tree before the move.
If \isa{c} did not exist in the tree, \isa{oldp} is set to \isa{None}.
Otherwise, \isa{oldp} records the previous parent and metadata of \isa{c}: if there exist \isa{p'} and \isa{m'} such that \isa{(p', m', c) {\isasymin} tree}, then \isa{oldp} is set to \isa{Some (p', m')}.
The \isa{get\_parent} function implements this (lines 16--20).

In the first line of \isa{get\_parent}, the expression between \isa{::} and \textbf{where} is the type signature of the function, in this case:
\begin{quote}
\isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set {\isasymRightarrow} 'n {\isasymRightarrow} ('n {\isasymtimes} 'm) option}
\end{quote}
This signature denotes a function that takes two arguments: a tree \isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set} and a node ID \isa{'n}.
It then returns a \isa{('n {\isasymtimes} 'm) option}.
The operator \isa{{\isasymexists}!x} means ``there exists a unique value \isa{x} such that\dots'', while \isa{THE x} means ``\emph{choose} the unique value \isa{x} such that\ldots''.

@@

The \isa{apply\_op} function (lines 42--49) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
The constraint \isa{'t::\{linorder\}} in the type signature indicates that timestamps \isa{'t} are instances of \isa{linorder} type class, and they can therefore be compared with the $<$ operator defining a linear (or total) order.
This comparison occurs on line 47.

@@

Due to the undo-do-redo cycle, the state of the tree is as if all operations had been applied using \isa{do\_op} in increasing timestamp order, regardless of the order in which they were actually applied.
This provides a clear and consistent approach to the handling of conflicts:
\begin{itemize}
    \item If two operations concurrently move the same node, the operation with the lower timestamp moves the node first, and then the operation with the greater timestamp moves it again, so the final parent is determined by the latter.
        Since move operations do not specify the old location of a node, but only the new location, this sequential execution of concurrent operations is well-defined.
    \item If two operations would introduce a cycle when combined, as in \S\ref{sec:move-cycle}, then the operation with the greater timestamp is ignored.
        This is the case because \isa{do\_op} checks for cycles based on the tree created by all operations with a lower timestamp.
        The lower of two conflicting operations will take effect, since that operation by itself is safe.
        When the higher-timestamped operation is applied, \isa{do\_op} detects that it would introduce a cycle, and therefore ignores the operation.
\end{itemize}

This resolution of a conflict between two operations can be generalised to any number of conflicting operations by applying the rules repeatedly pairwise.

@@

@@

This approach to determining ordering also easily supports reordering of child nodes within a parent: to move a node to a different position in a list, we use the list CRDT to generate a new ID at the desired position in the sequence~\cite{Kleppmann:2020jw}.
Then we perform a move operation in which the parent node is unchanged, but the metadata is changed to this new ID.

\smallbreak\noindent\textbf{File merging.}
In a distributed filesystem, replication and conflict resolution is required not only for the directory structure, but also for the contents of individual files.
This can be accomplished by using CRDTs for file contents as well.
We discuss distributed filesystems further in \S\ref{sec:filesystems}.

@@

A tree is an acyclic graph in which every node has exactly one parent, except for the root, which has no parent.
In fact, we slightly generalise this property and allow more than one root to exist, so the graph represents a forest, allowing an application to move nodes between different trees, if desired.
For example, the ``trash'' node used for deletion (\S\ref{sec:ops-trees}) can be separate from the main tree.
To prove that our algorithm maintains a forest structure, no matter which operations are applied, we demonstrate several invariants.

@@

\smallbreak\noindent\textbf{Other correctness properties}
There are further criteria we might use to determine if our algorithm is correct.
For example, we might demonstrate that a single-replica system operates with the usual sequential semantics of a tree.
In a system with multiple disjoint trees, we could prove that the trees don't get tangled together.
We conjecture that those properties hold for our algorithm, but leave the proof out of scope for this paper.

@@

@@

We used Isabelle/HOL's code generation mechanism to extract executable Scala code from our HOL definitions, as discussed in \S\ref{subsect.extracting}, and wrapped this implementation in a simple network service.
We deployed three replicas of this service on Amazon EC2 \texttt{c5.large} instances in different regions worldwide: in Northern California (region \texttt{us-west-1}), Ireland (\texttt{eu-west-1}), and Singapore (\texttt{ap-southeast-1}).
The network delays between these regions are substantial: in our experiments, a request-response round-trip from Ireland to California takes a median of 145~ms, while a request from Singapore to California takes 176~ms.

We use a synthetic workload in which each replica starts with an empty tree, and then generates move operations at a fixed rate; for each move operation the generating replica chooses parent and child nodes uniformly at random from a set of 1,000 tree nodes.
A tree node is created by the first operation that refers to it.
We use Lamport timestamps~\cite{Lamport:1978jq} as operation timestamps, and 64-bit integers as tree node identifiers.
When a replica generates an operation, it immediately applies that operation to its local state, and asynchronously sends the operation to the other two replicas via TCP connections.
When a replica receives an operation from a remote replica, it also applies that operation to its own replica state.
Thus, of the operations applied by each replica, approximately one third are locally generated, and two thirds are received from the other two replicas.

>>>>

\begin{document}
\title{A highly-available move operation for replicated trees}

@@

In distributed filesystems and collaborative multi-user software, this tree is replicated across multiple nodes.
If users attempt to update the tree concurrently on different replicas, concurrency control is required.
However, standard techniques such as two-phase locking require synchronous coordination between replicas.
If the software is running on mobile devices with unreliable network connectivity, an application based on synchronous coordination becomes unresponsive during network interruptions, leaving users unable to work while they are offline.

If we want to allow offline work, we must allow the system to continue processing read and write requests even in the presence of arbitrary network partitions; in other words, we require high \emph{availability} and \emph{partition-tolerance} in the sense of the CAP theorem~\cite{Gilbert:2002il}.
We can achieve this goal by using \emph{optimistic replication}~\cite{Saito:2005jw}, which means that any replica can make changes to the data without waiting for communication with any other replicas; updates made while disconnected are sent to other replicas later when a network connection is available.
Besides allowing the system to tolerate network partitions, this approach can also improve performance for end users because the response time of a user request is independent of any network latency.

@@

Our algorithm supports optimistic replication, allowing replicas to temporarily diverge as they are updated, and ensuring that they always converge towards a consistent state.
It is an example of a \emph{Conflict-free Replicated Data Type} or CRDT~\cite{Shapiro:2011un}, and it guarantees a consistency model called \emph{strong eventual consistency}~\cite{Shapiro:2011un,Gomes:2017gy}.
Our contributions are:
\begin{itemize}
    \item We define a Conflict-free Replicated Data Type for trees that allow \emph{move} operations without any coordination between replicas such as locking or consensus.
        As discussed in \S\ref{sec:impossibility}, this has previously been thought to be impossible to achieve \cite{Najafzadeh:2017vk,Najafzadeh:2018bw}.
    \item We formalise the algorithm using Isabelle/HOL \cite{DBLP:conf/tphol/WenzelPN08}, a proof assistant based on higher-order logic, and obtain a computer-checked proof of correctness.
        In particular, we prove that arbitrary concurrent modifications to the tree can be merged such that all replicas converge to a consistent state, while preserving the tree structure.
        Our proof technique can also be applied to other distributed systems, making it of independent interest.
    \item To demonstrate its practical viability, we extract a formally verified Scala implementation of our algorithm from Isabelle.
        We compare its performance to a hand-optimised (not formally verified) implementation and to classic state machine replication.
        In a geo-replicated setup across three continents, our algorithm has approximately half the throughput of state machine replication, but up to five orders of magnitude lower latency.
    \item We perform experiments with Dropbox and Google Drive, and show that they exhibit problems that would be prevented by our algorithm.
\end{itemize}

@@

On a filesystem, the destination directory of a move operation must not be a subdirectory of the directory being moved.
For example, if \texttt{b} is a subdirectory of \texttt{a}, then the Unix shell command \texttt{mv a a/b/} will fail with an error.
This restriction is required because allowing this operation would introduce a cycle into the directory graph, and so the filesystem would no longer be a tree.
Any tree data structure that supports a move operation must handle this case.

@@

Another possible outcome is shown in Figure~\ref{fig:move-cycle}b: the nodes involved in the concurrent moves (and their children) could be duplicated, so that both ``$A$ as a child of $B$'' and ``$B$ as a child of $A$'' can exist in the tree.
However, such duplication is undesirable for the same reasons as in \S\ref{sec:move-same-item}.

@@

\section{The replicated tree algorithm}\label{sec:algorithm}

@@

When the user wants to make a change to the tree, they generate an operation and apply it to their local replica.
Every operation is also asynchronously sent over the network to all other replicas, and applied by every remote replica using the same algorithm as for local operations.
A replica may communicate directly with any other replica.
The network may arbitrarily delay or reorder messages, but we assume that an underlying network protocol detects and retransmits lost messages, and suppresses duplicates.
We do not assume any central server or consensus protocol.
Any number of replicas may fail by crashing, and any non-crashed subset of replicas can continue executing operations.

The key consistency property of our algorithm is \emph{convergence}: that is, whenever any two replicas have applied the same set of operations, then they must be in the same state---even if the operations were applied in a different order on different replicas.
We prove this in \S\ref{sec:proof} by showing that applying operations using our algorithm is commutative.

@@

We allow the tree to be updated in three ways: by creating a new child of any parent node, by deleting a node, or by moving a node to be a child of a new parent.
However, all three types of update can be represented by a move operation.
To create a node, we generate a fresh ID for that node, and issue an operation to move this new ID to be a child of its desired parent; the node is then implicitly created.
We also designate as ``trash'' some node ID that does not exist in the tree; then we can delete a node by moving it to be a child of the trash.
Node creation and deletion are discussed further in \S\ref{sec:create}.

@@

When a replica applies a \isa{Move} operation to its tree, it also records a corresponding \isa{LogMove} operation in its log.
The \isa{t}, \isa{p}, \isa{m} and \isa{c} fields are taken directly from the \isa{Move} record, while the \isa{oldp} field is filled in based on the state of the tree before the move.
If \isa{c} did not exist in the tree, \isa{oldp} is set to \isa{None}.
Otherwise, \isa{oldp} records the previous parent and metadata of \isa{c}: if there exist \isa{p'} and \isa{m'} such that \isa{(p', m', c) {\isasymin} tree}, then \isa{oldp} is set to \isa{Some (p', m')}.

The \isa{get\_parent} function (lines 16--20) implements this.
In the first line of \isa{get\_parent}, the expression between \isa{::} and \textbf{where} is the type signature of the function, in this case:
\begin{quote}
\isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set {\isasymRightarrow} 'n {\isasymRightarrow} ('n {\isasymtimes} 'm) option}
\end{quote}
This signature denotes a function that takes two arguments: a tree \isa{('n {\isasymtimes} 'm {\isasymtimes} 'n) set} and a node ID \isa{'n}.
It then returns a \isa{('n {\isasymtimes} 'm) option}.
The operator \isa{{\isasymexists}!x} means ``there exists a unique value \isa{x} such that\dots'', while \isa{THE x} means ``\emph{choose} the unique value \isa{x} such that\ldots''.

@@

The \isa{apply\_op} function (lines 42--49) takes two arguments: a \isa{Move} operation to apply and the current replica state; and it returns the new replica state.
The constraint \isa{'t::\{linorder\}} in the type signature indicates that timestamps \isa{'t} are instances of the \isa{linorder} type class, and they can therefore be compared with the $<$ operator defining a linear (or total) order.
This comparison occurs on line 47.

@@

Due to the undo-do-redo cycle, the state of the tree is as if all operations had been applied using \isa{do\_op} in increasing timestamp order, regardless of the order in which they were actually applied.
This provides a clear and consistent approach to the handling of conflicts:
\begin{itemize}
    \item If two operations concurrently move the same node, the operation with the lower timestamp moves the node first, and then the operation with the greater timestamp moves it again, so the final parent is determined by the latter.
        Since move operations do not specify the old location of a node, but only the new location, this sequential execution of concurrent operations is well-defined.
        In Figure~\ref{fig:move-same-item}, the outcome is (c) if the operation from replica 1 has the greater timestamp, or (d) if the operation from replica 2 has the greater timestamp.
    \item If two operations would introduce a cycle when combined, as in \S\ref{sec:move-cycle}, then the operation with the greater timestamp is ignored because \isa{do\_op} checks for cycles based on the tree created by all operations with a lower timestamp.
        The lower of two conflicting operations will take effect, since that operation by itself is safe.
        When the higher-timestamped operation is applied, \isa{do\_op} detects that it would introduce a cycle, and therefore ignores the operation.
        In the example of Figure~\ref{fig:move-cycle}, the outcome is (c) if the operation from replica 1 has the lower timestamp, or (d) if the operation from replica 2 has the lower timestamp.
\end{itemize}

This resolution of a conflict between two operations can be generalised to any number of conflicting operations by repeatedly applying the rules pairwise.

@@

\subsection{Node creation and deletion}\label{sec:create}

As discussed previously, no separate operations for node creation and deletion are needed, since a node is implicitly created when it is first moved, and a node can be deleted by moving it to a designated trash node outside of the tree.
By avoiding a distinction between different operation types we simplify the algorithm and proofs of correctness, but we potentially sacrifice performance in applications that mostly create and delete nodes, and only occasionally move nodes.
Ideally, we would want to incur the costs of the undo-do-redo process only for genuine move operations, and not for node creation and deletion.

This optimisation is possible for operations that create new tree nodes: such operations can be directly applied to the tree without any undo/redo by using \isa{do\_op} instead of \isa{apply\_op}, and they do not need to recorded in the log.
We have proved in Isabelle that this optimisation is safe, under the following additional assumptions: 1.~the parent node in any move/create operation must exist in the tree; 2.~the creation operation for a given node must be unique (i.e.\ there cannot be two operations that both create the same node); 3.~a node creation operation is applied before any operation that moves the created node.
These assumptions are easily satisfied in practice.

For node deletion we have not been able to find a similar optimisation.
Deletion operations must go through the undo-do-redo process because deleting a node may cause a previously unsafe move operation with a greater timestamp to become safe by breaking a particular ancestor relationship; and as the previously ignored operation takes effect, we must re-evaluate all operations with greater timestamps to determine whether they are safe or not.

When a node is deleted, any children of the deleted node remain in the tree, but since they are no longer reachable from the root, they effectively become invisible to the application.
We do not recursively remove children, since a move operation concurrent to the deletion operation might move the deleted subtree back out of the trash and into the visible tree; in this scenario we want all children of the moved subtree to be preserved unmodified.

@@

This approach to determining ordering also easily supports reordering of child nodes within a parent: to move a node to a different position in a list, we use the list CRDT to generate a new ID at the desired position in the sequence~\cite{Kleppmann:2020jw}.
Then we perform a move operation in which the parent node is unchanged, and this new ID is used as metadata.

%\smallbreak\noindent\textbf{File merging.}
%In a distributed filesystem, replication and conflict resolution is required not only for the directory structure, but also for the contents of individual files.
%This can be accomplished by using CRDTs for file contents as well.
%We discuss distributed filesystems further in \S\ref{sec:filesystems}.

@@

A tree is an acyclic graph in which every node has exactly one parent, except for the root, which has no parent.
In fact, we slightly generalise this property and allow more than one root to exist, so the graph represents a forest, allowing an application to move nodes between different trees if desired.
For example, the trash node used for deletion can be separate from the main tree.
To prove that our algorithm maintains a forest structure, no matter which operations are applied, we demonstrate several invariants.

@@

\smallbreak\noindent\textbf{Other correctness properties.}
There are further criteria we might use to determine if our algorithm is correct.
For example, we might demonstrate that a single-replica system operates with the usual sequential semantics of a tree.
In a system with multiple disjoint trees, we could prove that the trees don't get tangled together.
We conjecture that those properties hold for our algorithm, but leave the proof out of scope for this paper.

@@

\begin{figure}
  \begin{center}
  \includegraphics{../evaluation/data/crdt-generated.pdf}
  \end{center}
  \caption{Median execution time to apply an operation to the replica state, using Scala code generated by Isabelle. Error bars indicate the minimum and 95th percentile.}
  \label{fig:plot-generated}
\end{figure}

\begin{figure}
  \begin{center}
  \includegraphics{../evaluation/data/crdt-optimised.pdf}
  \end{center}
  \caption{Median execution time to apply an operation to the replica state, using a hand-optimised (not formally verified) Scala implementation. Error bars indicate the minimum and 95th percentile.}
  \label{fig:plot-optimised}
\end{figure}

@@

Our algorithm can be computationally expensive if a large number of undos and redos are required.
To demonstrate that this cost is acceptable in practice, we evaluated the performance of two different implementations of our algorithm.
The first implementation is Scala code extracted from our formally verified HOL definitions using Isabelle/HOL's code generation mechanism, as discussed in \S\ref{subsect.extracting}.
The second implementation is handwritten Scala code that we believe to be functionally equivalent, but which we have not formally verified.
By comparing these implementations we can distinguish between inefficiencies introduced by code generation and costs that are inherent to our algorithm.

We wrapped both implementations in a simple network service and deployed three replicas of this service on Amazon EC2 \texttt{c5.large} instances in different regions worldwide: in Northern California (region \texttt{us-west-1}), Ireland (\texttt{eu-west-1}), and Singapore (\texttt{ap-southeast-1}).
The network delays between these regions are substantial: in our experiments, a request-response round-trip from Ireland to California takes a median of 145~ms, while a request from Singapore to California takes 176~ms.

\begin{figure*}
  \begin{center}
  \includegraphics{../evaluation/data/leader-vs-crdt.pdf}
  \end{center}
  \caption{Median time to apply a local CRDT operation compared to the median time to perform a move operation using state machine replication, with a leader located in another region. Note the log scale on both axes.}
  \label{fig:leader-plot}
\end{figure*}
